---
title: "DLM"
author: "Sky Yun"
date: "2026-04-16"
output: html_document
---

# Library Imports
```{r}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(forecast))
source("source/data_sources_and_dependencies.R")
source("scripts/mcmc_sampling_functions.R")
```

# MCMC Approximation
```{r}
n_samples = 10000
train_data=list(
  N = length(diff_y_scaled),
  y = diff_y_scaled
)

ma3_mod <- cmdstan_model("source/ma3.stan");
```

## MCMC Sampling
```{r}
ma3_fit <- ma3_mod$sample(
  seed=2026,
  chains=2, 
  refresh=n_samples,
  iter_warmup=2500, # Ensures that chain is sufficiently burned-in
  iter_sampling=n_samples,
  data=train_data, 
  show_exceptions=FALSE
)
```

### Summary
```{r}
ma3_mcmc_fit_summary <- ma3_fit$summary()

variables = c("mu","theta[1]", "theta[2]", "theta[3]", "sigma")
ma3_mcmc_clean_summary <- clean_mcmc_summary(variables, ma3_mcmc_fit_summary)

head(ma3_mcmc_fit_summary)
```

## MCMC Diagnostics

### Trace and Rank Plots
```{r}
# Samples
posterior_samples <- ma3_fit$draws(variables=c("mu", "sigma", "theta[1]"))

# Trace Plots
mcmc_trace(posterior_samples, facet_args=list(ncol=1)) + theme_minimal()

# Rank Plots
mcmc_rank_hist(posterior_samples) + theme_minimal()
```


### Posterior Predictive Checks

**Posterior Predictive Samples vs Observed Samples**
```{r}
# Checking Posterior Predictive Samples against Observed
y_preds_MCMC <- ma3_fit$draws(c("y_post_pred"), format="draws_matrix")
pp_check(diff_y_scaled, y_preds_MCMC[400:600, ], fun="dens_overlay")
```

**Cross-Validation: Calibration of Credible Intervals**

**MA(3) Cross-Validation Function**
```{r}
# Leave One out Method for 80% Credible Interval

ma3_mod_cv <- cmdstan_model("source/ma3_cross_val.stan")
# t_lower >= 4
# t_upper <= N
ci_ma3_cv <- function(mod, data, n_samples, lower_t, upper_t) {
  ci_lower = c()
  ci_upper = c()
  
  for (t in lower_t:upper_t) {
    train_data = list(
      N = length(data[-t]),
      y = data[-t],
      t_pred = t
    )
    
    mod_fit <- mod$sample(
      chains=1,
      iter_warmup = 2500,
      refresh = n_samples ,
      iter_sampling = n_samples,
      data=train_data,
      show_messages=FALSE,
      show_exceptions = FALSE)
  
      y_pred_samples <- as.vector(
        mod_fit$draws(variables=c("y_pred"), 
                      format="matrix")
        )
      ci = quantile(y_pred_samples, c(0.1, 0.9))
      ci_lower = c(ci_lower, ci["10%"])
      ci_upper = c(ci_upper, ci["90%"])
      
      print(paste("Done CI for time:", t))
  }
  return(data.frame(ci_lower, ci_upper))
}
```

**Running Credible Intervals**
```{r}
# If Slow at start rerun function
l_t = 5
u_t = 79
ci_ma3 <- ci_ma3_cv(ma3_mod_cv, diff_y_scaled, n_samples=5000, lower_t=l_t, upper_t=u_t)
```

**Cross-Validation: Calibration of Credible Intervals**
```{r}
ci_and_obs = ci_ma3
ci_and_obs['obs_diff_y'] = diff_y_scaled[l_t:u_t]
ci_and_obs <- ci_and_obs %>% 
              mutate(is_within_ci = (obs_diff_y >= ci_lower) & 
                                    (obs_diff_y <= ci_upper))
ci_and_obs["time"] = l_t:u_t
sum(ci_and_obs$is_within_ci) / nrow(ci_and_obs)
```

**Cross-Validation Visualized**
```{r}
ci_cv_plot <- ggplot(ci_and_obs, aes(x=time)) +
    geom_errorbar(aes(ymin=ci_lower, ymax=ci_upper, color=is_within_ci)) +
    geom_point(aes(y=obs_diff_y, color=is_within_ci)) +
    xlab("Time Index") +
    ylab("Observed diff") +
    labs(title="Credible Intervals Cross-Validation") +
    theme(legend.position="none")
ci_cv_plot
```


## Assessing Predictions 1-Step Forecast

**MCMC Sampling for Predicition of One-Step Forecast**
```{r}
N = length(diff_y_scaled)

train_data = list(
  N = N-1,
  y = diff_y_scaled[-N],
  t_pred = N
)

ma3_pred_fit = ma3_mod_cv$sample(
      chains=1,
      seed = 2027,
      iter_warmup = 2500,
      refresh = 10000 ,
      iter_sampling = 10000,
      data=train_data,
      show_messages=FALSE,
      show_exceptions = FALSE)
```

**Compare test with prediction samples**
```{r}
source("scripts/one_step_prediction_functions.R")

y_pred_samples = ma3_pred_fit$draws(variable="y_pred", format="df")
y_pred_samples["diff_y_pred"] = y_pred_samples["y_pred"]
y_pred_samples["y_pred"] = y_pred_samples["diff_y_pred"] + y_scaled[N]

y_pred = y_pred_samples$y_pred
test_point = diff_y_scaled[N] + y_scaled[N]   
y_pred_median = median(as.numeric(y_pred_samples$y_pred))

# Density Plot
summary_one_step <- summary_one_step(y_pred_samples, test_point)
plot_one_step_forecast_density(y_pred_samples, test_point)
summary_one_step
```


# Writing CSV Files for Report
```{r}
# write.csv(summary_one_step, "figs/tables/03_ma3_mcmc_pred_summary.csv", row.names=FALSE)
# write.csv(ma3_mcmc_clean_summary, "figs/tables/03_ma3_mcmc_clean_summary.csv")
```

