---
title: "MCMC"
author: "Sky Yun"
date: "2026-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(bayesplot))
source("source/data_sources_and_dependencies.R")
```
# Plots
```{r}
plot(dataset$MSPUS)
```


# Simple BLR MCMC Approximation

## MCMC Sampling
```{r}

train_data=list(
  N = nrow(dataset),
  y = dataset$MSPUS,
  time = dataset$time_index,
  gdp = dataset$GDP,
  unempl_rate = dataset$unemployment_rate,
  mortgage_rate = dataset$mortgage_rate
)
n_samples = 5000

blr_mod <- cmdstan_model("bayesian_regression.stan")

blr_fit <- blr_mod$sample(
  chains=2,
  refresh=n_samples,
  iter_warmup=1000,
  iter_sampling=n_samples,
  data=train_data
)
```
### Summary

```{r}
blr_fit$summary()
```


## MCMC Diagnostics

### Trace Plots and Rank Plots: Fast-Mixing Conditions
```{r}
posterior_samples <- blr_fit$draws(variables=c("beta0", "beta1", "sigma"), format="draws_df")

# Trace Plots
mcmc_trace(posterior_samples, facet_args=list(ncol=1)) + theme_minimal()

# Rank Plots
mcmc_rank_hist(posterior_samples) + theme_minimal()
```
- Trace plots samples oscillate rapidly indicating fast-mixing 
- Rank plots are approximately uniform which also indicates fast-mixing conditions.

### Posterior Predictive Checks: Calibration

**Posterior Predictive Plots**:
```{r}
# Check First 100 Reps
y_preds_MCMC <- blr_fit$draws(c("y_pred"), format="draws_matrix")[1:100, ]
pp_check(dataset$MSPUS, y_preds_MCMC, fun="dens_overlay")
```
- Fits pretty well for the first peak, might be underfitting for the second peak.  

**Cross-Validation: Prediction Accuracy using MAE and Cross-Validation Prediction Intervals**














