---
title: "DLM"
author: "Sky Yun"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Library Imports
```{r}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(forecast))
source("source/data_sources_and_dependencies.R")
```


## Time Series Exploratory Analysis

**Time Series Plot**
```{r}
data=data_scaled
plot(x=data$DATE, y=data$MSPUS, 
     ylab="Median US House Price", 
     xlab="Date", type="l")
```

- shows a slight increasing trend with slight fluctuations throughout time, potentially suggesting a Bayesian dynamic linear model (DLM)


**ACF and PACF plots of Differenced Series**
```{r}
# Predictions
diff_y = diff(dataset$MSPUS)
plot(diff_y, type="l")
```


```{r}

# Find best ARIMA model based on AIC
arima_best_fit = auto.arima(diff_y)
arima_1_1 = arima(diff_y, order=c(1, 1, 1))

# mu, ar1, ma1, ma2 parameters
summary(arima_best_fit)

# mu, ar1, ma1 parameters
summary(arima_1_1)
```

**Residual Checking**
```{r}
# Approximately Normal Residuals
# MA(3) Model
qqnorm(arima_best_fit$residuals, main="MA(3)")
qqline(arima_best_fit$residuals)

# ARIMA(1, 1) Model
qqnorm(arima_1_1$residuals, main="ARMA(1, 1)")
qqline(arima_1_1$residuals)
```

## MCMC Approximation

**Compiling Model**
```{r}
train_data=list(
  N = length(diff_y),
  y = diff_y
)

n_samples = 10000

ma3_mod <- cmdstan_model("source/ma3.stan");
# arma11_mod <- cmdstan_model("source/arima.stan");
```

**Sampling**
```{r}
dlm_fit <- dlm_mod$sample(
  seed=2,
  chains=2, 
  refresh=n_samples,
  iter_warmup=5000,
  iter_sampling=n_samples,
  data=train_data,
  show_exceptions = FALSE
)
```

## MCMC Diagnostics

### Summary
```{r}
dlm_fit$summary()
```



### Trace and Rank Plots
```{r}
# Samples
posterior_samples <- dlm_fit$draws(variables=c("mu", "theta"))

# Trace Plots
mcmc_trace(posterior_samples, facet_args=list(ncol=1)) + theme_minimal()

# Rank Plots
mcmc_rank_hist(posterior_samples) + theme_minimal()
```


**T**
```{r}
y_preds_MCMC <- dlm_fit$draws(c("y_post_pred"), format="draws_matrix")[1:50,]
pp_check(diff_y, y_preds_MCMC, fun="dens_overlay")
```


**Code**
```{r}
ggplot(as.data.frame(log_y), aes(x=log_y)) + geom_density()

vec_mcmc = as.vector((y_preds_MCMC[7, ]))
y_pred_mcmc = as.data.frame(x = vec_mcmc)
colnames(y_pred_mcmc) = c("y_pred")

pp_check(vec_mcmc, y_preds_MCMC, fun="dens_overlay")
pp_check(log_y, y_preds_MCMC, fun="dens_overlay")

plot(y=vec_mcmc, x=dataset$time_index, type="l")
plot(y=log_y, x=dataset$time_inde, type="l")
```





