---
title: "STAT405 Proposal"
output: pdf_document
author: "Lily Wang (00000000), Sky Yun (55333710), Zhengxi Yuan (00000000)"
date: "2026-03-09"
urlcolor: blue
---

**Repository Link**: https://github.com/yun-sky/STAT-405-Project

# Scientific Question
How well does a Bayesian linear regression model fitted on time predict quarterly median US house prices compared to another Bayesian Linear Regression model fitted on purely economic predictors (excluding time)?


```{r echo=FALSE}
suppressPackageStartupMessages(library(dplyr))

house_price <- read.csv("https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/MSPUS.csv", header=TRUE)
economic_factors <- read.csv("https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/US_House_Price.csv", header=TRUE)
california_price <- read.csv("https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/housing.csv", header=TRUE)
```

## Main Dataset(s)
**Economic Predictors for House Prices**: [Source](https://www.kaggle.com/datasets/jyotsnagurjar/factors-influencing-us-house-prices?resource=download) and [Direct Download](https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/US_House_Price.csv)

```{r}
head(economic_factors)
```

**Quartly Median US House Prices**: [Source](https://fred.stlouisfed.org/series/MSPUS) and [Direct Download](https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/MSPUS.csv)


```{r}
head(house_price)
```

## Back Up Dataset
**Median California Housing Prices**: [Source](https://www.kaggle.com/datasets/camnugent/california-housing-prices) and [Direct Download](https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/housing.csv)

- **Note:** Research question will change for a new set of predictors but methodology will be the same

```{r}
head(california_price)
```

# Methodology
1. Data Cleaning and Exploratory Analysis
   - Merge economic predictors data with median housing prices.
   - Determining the appropriate prior and likelihood for each BLR model. 
2. Model Fitting: Training BLR Models
   - Find the posteriors of each BLR model. Without conjugate priors we will need to estimate the posteriors using two different methods: SNIS or MCMC
3. Model-Checking and Diagnostics: Checking Prior and Likelihood
4. Model-Comparison: Using a Loss Function to Evaluate Model Performance 
   - We will predict the test data by generating point estimates from the mode of the posterior predictive     distribution. 
   - We then assess the performance of each model by comparing their RMSE values. 
5. Results
   - Determine the best model based on RMSE and Discuss the trade-offs of using one model over another to predict quarterly median house prices. 


# Group Contributions
Weekly meetings will be held to ensure that the project is on track. All members will contribute to group discussions and contribute to at least a part of each phase which roughly corresponds to stages in our methodology. Here is a rough timeline of when we will complete each phase. 

Phase 1 (March 20th): Exploratory Analysis and Data-Cleaning (Lily & Jason)

Phase 2 (March 27th): Models Fitting and Model-Checking (Sky, Jason, Lily)

Phase 3 (April 10th): Model Testing and Comparison (Sky, Lily, Jason)

Phase 4 (April 15th): Interpretation and Results (Sky)

Phase 5 (April 15th): Submission (Lily)



