---
title: "STAT405 Proposal"
output: pdf_document
author: "Lily Wang, Sky Yun, Zhengxi Yuan"
date: "2026-03-09"
---

**Repository Link**: https://github.com/yun-sky/STAT-405-Project

# Datasets

```{r echo=FALSE}
suppressPackageStartupMessages(library(dplyr))

house_price <- read.csv("https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/MSPUS.csv", header=TRUE)
economic_factors <- read.csv("https://raw.githubusercontent.com/yun-sky/STAT-405-Project/refs/heads/master/Datasets/US_House_Price.csv", header=TRUE)

# house_price <- rename(house_price, DATE = observation_date)
# dataset <- merge(house_price, economics_factors, by = "DATE")
```

## Main Dataset
**Economic Predictors for House Prices**

https://www.kaggle.com/datasets/jyotsnagurjar/factors-influencing-us-house-prices?resource=download 

```{r}
head(economic_predictors)
```

**Quartly Median US House Prices**

https://fred.stlouisfed.org/series/MSPUS

```{r}
head(house_price)
```

## Back Up Dataset




## Scientific Question
How well does a Bayesian linear regression model fitted on time predict quarterly median US house prices compared to another Bayesian Linear Regression model fitted on purely economic predictors (excluding time)?


## Methodology
1. Data Cleaning and Exploratory Analysis
   - Merge economic predictors data with median housing prices
2. Model Fitting: Training BLR Models
   - Find the posteriors of each BLR model. Without conjugate priors we will need to estimate the posteriors using two different methods: SNIS or MCMC
3. Model-Checking and Diagnostics: Checking Prior and Likelihood
4. Model-Comparison: Using a Loss Function to Evaluate Model Performance 
   - We will predict the test data by generating point estimates from the mode of the posterior predictive     distribution. 
   - We then assess the performance of each model by comparing their RMSE values. 
5. Results
   - Determine the best model based on RMSE and Discuss the trade-offs of using one model over another to predict quarterly median house prices. 


## Group Contributions
Weekly meetings will be held to ensure that the project is on track. All members will contribute to group discussions and contribute to each phase together. Here is a rough outline of who will do what. 

Phase 1: Exploratory Analysis (Lily & Jason)

Phase 2: Models Fitting (Sky, Lily, Jason)

Phase 3: Models Checking and Comparison (Sky, Lily, Jason)

Phase 4: Interpretation and Results (Sky, Lily, Jason)



