---
title: "STAT405 Final Report"
output: pdf_document
author: "Lily Wang (62800495), Sky Yun (55333710), Zhengxi Yuan (15025604)"
header-includes:
    - \usepackage{float}
date: "2026-03-23"
---

```{r, include = FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(ggplot2))
```

# **Introduction**

Our team worked with two U.S. housing-related datasets to study how well bayesian methods can predict housing prices in the US. The main goal of this project is to compare a Bayesian linear regression model with a time-series model, to determine which model better predicts quarterly median U.S. house prices.

The response dataset is the "Quarterly Median U.S. House Prices" dataset, and the explanatory dataset is the "Economic Predictors for House Prices" dataset, which consists monthly economic and housing-related factors. Based on our proposal, we decided to include the following variables:

-   `MSPUS` (Response variable): the quarterly median sales price of houses sold in the United States.
-   `GDP`: how the US economy impacts housing prices.
-   `mortgage_rate`: Percentage interest charged on a home loan.
-   `unemployment_rate`: Percentage of those unemployed in the US economy.

Since the response variable is recorded quarterly while the economic predictors are recorded monthly, the monthly predictor data must be aggregated to the quarterly level before merging with the quarterly housing price data. After merging the two data sets together, we checked for missing values and found no missing observations. Since we are doing a Bayesian analysis, it is hard to explore the posterior regions efficiently enough due to large scales. This analysis uses scaled values to make it easier to put priors on our model parameters.

```{r, echo=FALSE}
source("../source/data_sources_and_dependencies.R")
knitr::kable(head(dataset), caption = "US Median Housing Prices")
```

# **Literature Review**

Previous study on the Median US Housing Price written by Yuxiao Chen argued that during the federal reserve’s continuous interest rate increases in 2022-2023, the US house prices were expected to drop, but not fall sharply. The results showed that US house prices were still increasing, but at a lower rate. First, mortgage rates don't depend solely on interest rates, it is influenced more by inflation expectations and US federal demand for US bonds. Also, housing prices are sticky, so they may take years to adjust to market changes. 

This previous study focuses on economic argument with literature discussion, while our analysis is much more statistical and Bayesian. We build Bayesian models to predict and explain median US housing prices using macroeconomic factors, like GDP, unemployment rate, and mortgage rate. The study done by Yuxiao Chen relies more on economic reasoning and historical discussions. 

# **Simple Linear Regression Model**

Our naive model:

$$
y_t = \beta_0 + \beta_1 \, \text{time}_t + \beta_2 \, \text{GDP}_t + \beta_3 \, \text{UnemploymentRate}_t + \beta_4 \, \text{MortgageRate}_t + \varepsilon_t
$$

Our prior and likelihood:

$$
\begin{aligned}
\beta_t &\sim \mathcal{N}(0,10), \quad t \in \{0,1,2,3,4\} \\
\sigma  &\sim \text{Exponential}(1)
\end{aligned}
$$

$$
\begin{aligned}
y_t|\mu_t &\sim N(\mu_t, \sigma) \\
\mu_t &= \beta_0 + \beta_1 \,\text{time}_t + \beta_2 \,\text{GDP}_t + \beta_3 \,\text{UnemploymentRate}_t + \beta_4 \,\text{MortgageRate}_t
\end{aligned}
$$

### Simple Linear Regression Model with SNIS

After initializing and running the SNIS algorithm with re-sampling, we came to the following outputs. The calculated ESS and posterior mean for each parameter are:
```{r, echo = FALSE}
#setup
set.seed(405)

dataset <- dataset %>%
  na.omit() %>%
  mutate(
    MSPUS = as.numeric(scale(MSPUS)),
    time_index = as.numeric(scale(time_index)),
    GDP = as.numeric(scale(GDP)),
    unemployment_rate = as.numeric(scale(unemployment_rate)),
    mortgage_rate = as.numeric(scale(mortgage_rate))
  )


xs <- model.matrix(
  ~ time_index + GDP + unemployment_rate + mortgage_rate,
  data = dataset
)

y <- dataset$MSPUS
p <- ncol(xs)


#prior
beta_prior_sd <- 2.5
sigma_prior_rate <- 1

log_prior <- function(theta) {
  beta <- theta[1:p]
  sigma <- theta[p + 1]

  if (sigma <= 0) return(-Inf)

  sum(dnorm(beta, mean = 0, sd = beta_prior_sd, log = TRUE)) +
    dexp(sigma, rate = sigma_prior_rate, log = TRUE)
}


#likelihood
log_likelihood <- function(theta, xs, y) {
  beta <- theta[1:p]
  sigma <- theta[p + 1]

  if (sigma <= 0) return(-Inf)

  mu <- as.vector(xs %*% beta)
  sum(dnorm(y, mean = mu, sd = sigma, log = TRUE))
}


#proposal distribution
ols_fit <- lm(
  MSPUS ~ time_index + GDP + unemployment_rate + mortgage_rate,
  data = dataset
)

proposal_beta_mean <- coef(ols_fit)
proposal_beta_cov <- vcov(ols_fit)

proposal_log_sigma_mean <- log(summary(ols_fit)$sigma)
proposal_log_sigma_sd <- 0.35


#multivariate normal log-density
log_dmvnorm <- function(x, mean, sigma) {
  log_det <- as.numeric(determinant(sigma, logarithm = TRUE)$modulus)
  quad <- mahalanobis(x, center = mean, cov = sigma)
  -0.5 * (length(x) * log(2 * pi) + log_det + quad)
}


#proposal sampler and proposal density
sample_proposal <- function() {
  beta <- as.numeric(MASS::mvrnorm(1, mu = proposal_beta_mean, Sigma = proposal_beta_cov))
  sigma <- exp(rnorm(1, mean = proposal_log_sigma_mean, sd = proposal_log_sigma_sd))
  c(beta, sigma)
}

log_proposal <- function(theta) {
  beta <- theta[1:p]
  sigma <- theta[p + 1]

  if (sigma <= 0) return(-Inf)

  log_dmvnorm(beta, mean = proposal_beta_mean, sigma = proposal_beta_cov) +
    dlnorm(sigma, meanlog = proposal_log_sigma_mean, sdlog = proposal_log_sigma_sd, log = TRUE)
}


#normalize log-weights
normalize_log_weights <- function(log_w) {
  m <- max(log_w)
  w <- exp(log_w - m)
  w / sum(w)
}


#SNIS algorithm
snis <- function(N, xs, y) {
  draws <- matrix(NA, nrow = N, ncol = p + 1)
  colnames(draws) <- c(colnames(xs), "sigma")

  log_weights <- numeric(N)

  for (n in 1:N) {
    theta_n <- sample_proposal()
    draws[n, ] <- theta_n

    log_weights[n] <- log_likelihood(theta_n, xs, y) +
      log_prior(theta_n) -
      log_proposal(theta_n)
  }

  weights <- normalize_log_weights(log_weights)
  ess <- 1 / sum(weights^2)

  list(
    draws = draws,
    weights = weights,
    log_weights = log_weights,
    ess = ess
  )
}


#run SNIS
N <- 10000
posterior <- snis(N, xs, y)

posterior$ess


#posterior means
posterior_mean <- colSums(posterior$draws * posterior$weights)
posterior_mean
#resample approximate posterior draws
n_post <- 5000

posterior_index <- sample(
  1:nrow(posterior$draws),
  size = n_post,
  replace = TRUE,
  prob = posterior$weights
)
```

```{r, include = FALSE}
posterior_draws <- posterior$draws[posterior_index, , drop = FALSE]
head(posterior_draws)
```

Summary of the posterior results:
```{r, echo=FALSE}
set.seed(405)
#posterior summary table
posterior_summary <- data.frame(
  parameter = colnames(posterior_draws),
  mean = colMeans(posterior_draws),
  sd = apply(posterior_draws, 2, sd),
  q05 = apply(posterior_draws, 2, quantile, probs = 0.05),
  q50 = apply(posterior_draws, 2, quantile, probs = 0.50),
  q95 = apply(posterior_draws, 2, quantile, probs = 0.95)
)

post_sum <- posterior_summary |>
  select(-q05, -q50, -q95)

knitr::kable(post_sum, caption = "BLR SNIS Posterior Summary Table")
```

```{r, include=FALSE}
#posterior predictive simulation
n_obs <- nrow(xs)
y_rep <- matrix(NA, nrow = n_obs, ncol = n_post)

for (s in 1:n_post) {
  beta_s <- posterior_draws[s, 1:p]
  sigma_s <- posterior_draws[s, p + 1]

  mu_s <- as.vector(xs %*% beta_s)
  y_rep[, s] <- rnorm(n_obs, mean = mu_s, sd = sigma_s)
}


#posterior predictive summaries
y_pred_mean <- rowMeans(y_rep)
y_pred_q05 <- apply(y_rep, 1, quantile, probs = 0.05)
y_pred_q95 <- apply(y_rep, 1, quantile, probs = 0.95)

prediction_summary <- data.frame(
  observed = y,
  pred_mean = y_pred_mean,
  pred_q05 = y_pred_q05,
  pred_q95 = y_pred_q95
)

head(prediction_summary)
```

Doing posterior predictive checks, we get the following:
```{r, echo = FALSE}
#posterior predictive checks
ppc_summary <- data.frame(
  statistic = c("mean", "sd"),
  observed = c(mean(y), sd(y)),
  replicated_mean = c(mean(colMeans(y_rep)), mean(apply(y_rep, 2, sd))),
  bayesian_p_value = c(
    mean(colMeans(y_rep) >= mean(y)),
    mean(apply(y_rep, 2, sd) >= sd(y))
  )
)
knitr::kable(ppc_summary, caption = "BLR SNIS Posterior Predictive Checks")
```

We fitted a Bayesian linear regression model to standardized house prices (MSPUS) using time_index, GDP, unemployment_rate, and mortgage_rate as predictors. Posterior inference was conducted using self-normalized importance sampling (SNIS).

The effective sample size of our SNIS approximation is 3067.872 out of 10,000 draws. This suggests that the proposal distribution matched the posterior, stating that the posterior approximation is stable.

Looking at the posterior mean coefficients, GDP seems to be the strongest predictor, with a value of 1.09. The other predictors show relatively small positive and negative effects on median US housing price. The posterior mean of the residual standard deviation is around 0.27, suggesting there are still some unexplained variation remaining.

Posterior predictive checks indicate that the model fits the data reasonably well. The replicated mean and observed mean are both close to 0 and the standard deviations are similar. Also, the Bayesian p-values are near 0.5. In addition, the predictive plots show that the model captures the overall upward trend in house prices.

Overall, this model provides a good Bayesian baseline for the house-price data. It yields stable posterior inference, identifies GDP as the main predictor, and captures the main pattern in the data.

## Simple Linear Regression Model with MCMC

```{r child="02_blr_mcmc.Rmd"}
```


## BLR SNIS VS. MCMC

We compared SNIS and MCMC for the same baseline Bayesian linear regression model. The two methods produced very similar posterior results. The posterior means for SNIS and MCMC were 0.0003 and -0.0003 for the intercept, -0.0419 and -0.0660 for time_index, 1.0866 and 1.11 for GDP, -0.0295 and -0.0233 for unemployment_rate, 0.1681 and 0.170 for mortgage_rate, 0.2715 and 0.271 for sigma. In both methods, GDP was the strongest positive predictor.

The posterior uncertainty estimates were also close. For example, the posterior standard deviation of GDP was about 0.298 under SNIS and 0.313 under MCMC, while the posterior standard deviation of sigma was about 0.0224 under SNIS and 0.0226 under MCMC. This suggests that SNIS approximated both the center and spread of the posterior well.

The predictive results were very similar. For the first observation, SNIS gave a predictive mean of about -1.635, while MCMC gave a predictive mean of about -1.640. Both methods also produced fitted values that followed the overall upward trend in house prices. Overall, both methods performed well and pretty similar.


# **Time-Series MA(3)**

Our second model is a time-series model that only depends on past-lagged values, capturing the dynamic structure of the data, including trends, seasonality, and time dependence. To find the best-fit model based on AIC, we ran `auto.arima()` and it suggested the moving Average model with three lagged forecast error terms (MA(3)) to be the best. We performed first-order differencing of the scaled median housing price to align with the stationary assumptions of moving-average models.

Assume that $N$ is the size of the observed data $y$. We define the following Bayesian MA(3) model, where the right side describes our transformed parameters and likelihood distribution and the left defines our priors. 

\begin{align*}
e_1 &= y_1 - \mu & \mu & \sim \mathrm{Normal}(0, 1)\\
e_2 &= y_2 - (\mu + \theta_1 e_1) & \theta_k & \sim \mathrm{Normal}(0, 0.5) \;\;\; \text{for $t=1, 2, 3$} \\
e_3 &= y_3 - (\mu + \theta_1 e_2 + \theta_2 e_1) & \sigma & \sim \mathrm{Cauchy(0, 1)}\\
e_t &= y_t - (\mu + \theta_1 e_{t-1} + \theta_2 e_{t-2} + \theta_3 e_{t-3}) \\
Y_t &| \sigma, \mu, e_{t-1}, e_{t-2}, e_{t-3} \sim \mathrm{Normal\left(\mu + \sum_{i = 1}^{3} \theta_i e_{t-i}, \sigma\right)} \;\;\; \text{for $t = 4, ..., N$}
\end{align*}

Priors (and associated hyper-parameters) were chosen to reflect the general range and scale of our parameters of interest. For example, $\sigma$ is only allowed to take positive, real numbers so exponential and half-normal distribution were used as priors initially but proved to be not informative enough for the MCMC  the converge reliably. Additionally, to ensure stationarity (refer to appendix for more information), $\theta$ values should be within the range $[-1, 1]$. So a normal distribution was chosen with a small standard deviation to reflect the symmetric bounds around 0.

## MA(3) with MCMC

```{r child="03_ma3_mcmc.Rmd"}
```

## MA(3) with Variational Inference

The following table produces the VI posterior summaries for the model parameters. Each parameter is presented with its mean, median, standard deviation, median absolute deviation, and the 5th and 95th posterior quantiles.

```{r, include = FALSE}
suppressPackageStartupMessages(library(cmdstanr))
suppressPackageStartupMessages(library(bayesplot))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(posterior))

diff_y_scaled <- diff(data_scaled$MSPUS)


train_data <- list(
  N = length(diff_y_scaled),
  y = diff_y_scaled
)

ma3_stan <- cmdstan_model("../source/ma3.stan")

ma3_vi <- ma3_stan$variational(
  data = train_data,
  seed = 405,
  algorithm = "meanfield",
  iter = 10000,
  tol_rel_obj = 0.001,
  draws = 10000
)

params_of_interest <- c("mu", "theta[1]", "theta[2]", "theta[3]", "sigma")

vi_summary <- posterior::as_draws_df(ma3_vi$draws()) %>%
  dplyr::select(any_of(params_of_interest)) %>%
  tidyr::pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  dplyr::group_by(variable) %>%
  dplyr::summarise(
    mean   = mean(value),
    median = median(value),
    sd     = sd(value),
    mad    = mad(value),
    q5     = quantile(value, 0.05),
    q95    = quantile(value, 0.95),
    .groups = "drop"
  )
```

```{r, echo = FALSE}
knitr::kable(vi_summary, caption = "MA(3) VI Posterior Summary Table")
```

The intercept $\mu$ (mean = 0.0569, 90% CI: 0.020–0.093) captures a small upward drift in quarterly US house price changes. The entire credible interval above zero states that this trend is not just white noise.

$\theta_3$ from the model has the largest positive mean value (mean = 0.322, 90% CI: 0.144–0.487), which indicates that shocks from three quarters ago pushes quarterly house price in the same direction. $\theta_2$ shows a weaker effect at 0.175, while $\theta_1$ suggests negative effect with a value of -0.150. The noise term $\sigma$ (mean = 0.150, 90% CI: 0.131–0.172) is not too large, indicating the model is fairly stable.

### One-Step Forecast
Similar to before, we randomly sample from the posterior distribution for the one-step forecast using VI method. Below are sample distribution and prediction summaries.

```{r, include = FALSE}
source("../scripts/one_step_prediction_functions.R")
ma3_mod_cv <- cmdstan_model("../source/ma3_cross_val.stan")

N <- length(diff_y_scaled)

train_data_pred <- list(
  N      = N - 1,
  y      = diff_y_scaled[-N],
  t_pred = N - 1        
)

ma3_vi_pred <- ma3_mod_cv$variational(
  data        = train_data_pred,
  seed        = 405,
  algorithm   = "meanfield",
  iter        = 50000,
  tol_rel_obj = 0.001,
  draws       = 10000
)

y_pred_samples <- as.vector(
  ma3_vi_pred$draws(variables = "y_pred", format = "matrix")
)

test_point <- diff_y_scaled[N]

pred_summary <- data.frame(
  mean   = mean(y_pred_samples),
  median = median(y_pred_samples),
  sd     = sd(y_pred_samples),
  q5     = quantile(y_pred_samples, 0.05),
  q95    = quantile(y_pred_samples, 0.95),
  actual = test_point
)
print(pred_summary)
```

```{r, echo=FALSE, fig.show='hold', out.width='50%'}
plot_one_step_forecast_density(
  data.frame(y_pred = y_pred_samples), 
  test_point
)

mase_vi <- compute_mase(
  data.frame(y_pred = y_pred_samples), 
  test_point
)
mase_vi
```

The sample distribution is roughly normal, with the test point being very close to the posterior predictive mean. This suggests that the samples well approximate the one-step forecasts. Our VI MASE value of 0.80 is very close to our MCMC method, indicating that the two methods provide nearly identical accuracy.

## Time-Series MCMC VS. VI
```{r, echo=FALSE}
knitr::kable(read.csv("../figs/tables/03_ma3_mcmc_clean_summary.csv", header=TRUE), caption = "MA(3) MCMC Posterior Summary Table")

knitr::kable(vi_summary, caption = "MA(3) VI Posterior Summary Table")
```
Comparing the two posterior summary tables for the MA(3) model, the parameters estimates for the two methods are almost identical. Both methods indicate a negative first lag effect, and positive second and third lag effects, with $\theta_3$ being the largest. 

For forecasting performance, the two methods are essentially equivalent, with MCMC MASE value of 0.8104 and VI MASE value of 0.8010. This suggests that VI is a great approximation method, as it is much faster than the MCMC method, with minimal loss in predictive accuracy.

```{r child="05_conclusion.Rmd"}

```

# Appendix
```{r child="06_appendix.Rmd"}

```

